---
title: "Toronto Crime Rates"
subtitle: "Disadvantaged communities face continued marginalization."
author: "Rachael"
thanks: "Code and data are available at: https://github.com/rachaellam/Toronto-Crime-Rates.git."
date: "`r format(Sys.time(), '%d %B %Y')`"
abstract: "Neighbourhood Crime Rate data was pulled from the City of Toronto Open Portal. Although crime rates have remained consistent between 2014-2019, certain neighbourhoods experience higher crime rates than other in the Greater Toronto Area. Further analysis of neighbourhoods who experience higher crime rates are also economically disadvantaged communities. This data, without a further analysis and understanding of socioeconomic circumstances could lead to higher police surveillance and further unjust action towards marginalized communities"
output:
  bookdown::pdf_document2:
toc: FALSE
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(opendatatoronto)
library(tidyverse)
library(devtools)
library(dplyr)
library(reshape2)
library(ggplot2)
library(bookdown)

```

# Introduction




# Data

Our data is of penguins (Figure \@ref(fig:bills)).

```{r, include=FALSE}
raw_data <- read_csv(here::here("inputs/data/raw_data.csv"))

toronto_crime_highest <- raw_data %>%
  as_tibble() %>%
  select(Neighbourhood, Robbery_2014, Robbery_2015, Robbery_2016, Robbery_2017, Robbery_2018, Robbery_2019, Robbery_AVG) %>% # focusing on robbery. There are other crimes to choose from if desired.
  slice_max(Robbery_AVG, n = 10) %>% # choosing 10 neighbourhoods with the highest robbery average between 2014-2019
  select(Neighbourhood, Robbery_2014, Robbery_2015, Robbery_2016, Robbery_2017, Robbery_2018, Robbery_2019) %>% # removing Robber_AVG column
  melt(id = c("Neighbourhood"), #using `reshape2` to change dataframe to long-format data for plotting. kept Neighbourhood as the ID
       variable.name = "Year",
       value.name = "Number_of_Robberies")

toronto_crime_highest$Year <- as.character(toronto_crime_highest$Year) # turned column from factor into character to mutate below

toronto_crime_highest <-
  toronto_crime_highest %>% # mutated values to only be year rather than type_year
  mutate(Year = case_when(
    str_detect(Year, "Robbery_2014") ~ "2014",
    str_detect(Year, "Robbery_2015") ~ "2015",
    str_detect(Year, "Robbery_2016") ~ "2016",
    str_detect(Year, "Robbery_2017") ~ "2017",
    str_detect(Year, "Robbery_2018") ~ "2018",
    str_detect(Year, "Robbery_2019") ~ "2019",
    TRUE ~ Year
  ))
```


```{r bills, fig.cap="Bills of penguins", echo = FALSE, warning=FALSE}

toronto_crime_highest%>%
  ggplot(aes(x = Year, y = Number_of_Robberies, color = Neighbourhood)) +
  geom_smooth(aes(group = Neighbourhood), se = FALSE) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(color = "Neighbourhoods",
       x = "Date",
       y = "Number of Robberies",
       title = "Toronto Crime Rate",
       subtitle = "Number of robberies per year") +
  theme_minimal()
```

Talk more about it.

Also bills and their average (Figure \@ref(fig:billssssss)). (Notice how you can change the height and width so they don't take the whole page?)

```{r billssssss, fig.cap="More bills of penguins", echo = FALSE, fig.width=8, fig.height=4}

toronto_crime_pop_top <- raw_data %>%
  as_tibble() %>%
  select(Neighbourhood, Population) %>%
  slice_max(Population, n = 10)

toronto_crime_pop_top %>%
  ggplot(aes(x = Neighbourhood, y = Population, fill = Neighbourhood)) +
  geom_bar(stat = "identity") +
  ylab("proportion") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Talk way more about it. 



# Model

\begin{equation}
Pr(\theta | y) = \frac{Pr(y | \theta) Pr(\theta)}{Pr(y)}  (\#eq:bayes)
\end{equation}

Equation \@ref(eq:bayes) seems useful, eh?

Here's a dumb example of how to use some references: In paper we run our analysis in `R` [@citeR]. We also use the `tidyverse` which was written by @thereferencecanbewhatever If we were interested in baseball data then @citeLahman could be useful. 


# Results

# Discussion

## First discussion point

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this. 

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

# Appendix {-}

\newpage


# References


